---
title: "Co-occurrence with biotic factors"
author: "Mathieu Castelli & Nicolas Tromas"
date: "10/04/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(tidyverse); library(magrittr)
```

## Data setup
### Files import

```{r import, cache=TRUE,echo = FALSE, message=FALSE}
syn_dist <- read_tsv(file = "data/Distance_SYN.txt", col_names = TRUE) %>% 
   rename(OTU_1 = Nodes_syn1, OTU_2 = Nodes_syn2)

all_sparcc <- read_tsv("data/All_sparcc_corr.txt") %>% 
   filter(qval < .05) %>% 
   rename(OTU_1 = `OTU 1`, OTU_2 = `OTU 2`)

ref_syn <- read_tsv("data/liste_syn.txt", col_names = FALSE) %>% 
   pull(X1)
```
- syn_dist regroups all pairwise sequencing distances of all SYNECHO
- all_sparcc is the result of the SPARCC process to obtain trustable correlations between OTUs from the reads. We keep only correlations with p-value lower than .05
- ref_syn is the list of SYNECHO labels

### Extracting relevant data
We start by extracting from the SPARCC correlations, only the correlations that involve one or two SYNECHO.

```{r extract SYNECHO correlations}
syn_corr <- all_sparcc %>%
   filter(OTU_1 %in% ref_syn | OTU_2 %in% ref_syn)
```

And reversely, we save all non SYNECHO taxa from the SPARCC process.
```{r extract non syn taxa}
non_syn_taxa <- with(syn_corr, c(OTU_1, OTU_2)) %>%
   unique() %>%
   keep(~! . %in% ref_syn)
```
### Checks
We check for duplicate (permutted) pairs of SYNECHO: generate a min column with the smallest value, and max for the other value.
```{r check permutted pairs}
syn_corr %>% 
   mutate(min = pmin(OTU_1, OTU_2),
          max = pmax(OTU_1, OTU_2)) %>% 
   distinct(min, max) %>% 
   nrow() == nrow(syn_corr)
```
There's no permuttations. We can move on.

## Generation of data
This is the main data generation. It's an iteration over all non SYNECHO taxons.
We run it first then explain each section separately.

```{r main data generation , cache = TRUE }
min_correlated_taxa <- 7
pairs_data <- map_dfr(non_syn_taxa, ~{
   
   # For each non SYNECHO taxon it extracts the correlations to SYNECHOs from the
   # SPARCC data by matching on either OTU column then keeping the matched OTU 
   # and its correlation (to the taxon).
   
   taxon_correlated_syns <- syn_corr %>%
      filter(OTU_1 == .x | OTU_2 == .x) %>%
      mutate(OTU = if_else(OTU_1 == .x, OTU_2, OTU_1)) %>% 
      select(OTU, corr)
   
   if (nrow(taxon_correlated_syns) < min_correlated_taxa) return(NULL)   
   
   # From n elements there's n x (n-1) / 2 unique pairs. We use R's combn to 
   # generate them then do two table joins to re-integrate each element's info.
   syn_pairs <- combn(taxon_correlated_syns$OTU, m = 2) %>%
      t() %>% 
      as_tibble() %>%
      rename(OTU_1 = V1, OTU_2 = V2) %>% 
      left_join(taxon_correlated_syns, by = c("OTU_1" = "OTU")) %>% 
      left_join(taxon_correlated_syns, by = c("OTU_2" = "OTU"), suffix = c("_1", "_2"))
   
   # Again, because of pairs, we join in two times by first joining the distance
   # info reading the pairs in one order, than the other.
   # We also generate the absolute differences between the two correlations.
   
   pairs_dist_by_12 <- left_join(syn_pairs, syn_dist, by = c("OTU_1", "OTU_2"))
   
   pairs_dist_by_21 <- left_join(syn_pairs, syn_dist, by = c("OTU_2" = "OTU_1", "OTU_1" = "OTU_2"))
   
   pairs_dist_abs_corr_diff <-  pairs_dist_by_12 %>% 
      mutate(
         Distance = if_else(is.na(Distance),
                            pairs_dist_by_21$Distance,
                            Distance),
         Abs_diff = abs(corr_1 - corr_2)
      )
   res <- with(pairs_dist_abs_corr_diff, cor.test(Distance, Abs_diff))
   
   tibble(taxa = .x,
          nb_syns = nrow(taxon_correlated_syns),
          nb_pairs =  nrow(pairs_dist_abs_corr_diff),
          estimate = res$estimate,
          p_val = res$p.value,
          data = list(pairs_dist_abs_corr_diff))
}) %T>% 
   glimpse()
```

## Permutations & Bootstrapping
```{r permutations per non syn taxa, message=TRUE, warning=TRUE, cache=TRUE}
pairs_data_sign_corr <- pairs_data %>%
   filter(p_val < .05)

permutations_props <- map_dfr(seq_along(pairs_data_sign_corr$taxa), ~{
   browser()
   current <- pairs_data_sign_corr[., ]
   
   perm_cors <- rerun(1000,
                      current$data[[1]] %>% 
                         mutate(Distance = sample(Distance, replace = TRUE))
   ) %>% 
      map_dbl(~ cor(.$Abs_diff, .$Distance,method="spearman"))
   
   tibble(
      taxa = current$taxa,
      prop = (sum(abs(perm_cors) > abs(current$estimate)) + 1) / (length(perm_cors) + 1),
   )
})
pairs_data_perms_props <- pairs_data_sign_corr %>% left_join(permutations_props, by = c("taxa"))
```

## Analysis
### First look

```{r Means, cache = TRUE}
mean_corr_per_taxa <- pairs_data_perms_props$estimate %>% 
   abs() %>% 
   mean()

mean_corr_of_signif_levels <- pairs_data_perms_props %>% 
   filter(prop < .005) %>% 
   summarise(count = nrow(.),
             mean = mean(abs(estimate)))
```
The mean across all our non syn taxa sub-groups is `r mean_corr_per_taxa` (`r nrow(pairs_data_perms_props)` groups with at least seven SYNECHO correlated to the non syn taxa) and if we filter on significative based on our random permutations, the mean increases to `r mean_corr_of_signif_levels$mean` (with `r mean_corr_of_signif_levels$count` of groups that make the 0.005 cut).

Taking the absolute value of the significative correlations impacts a single taxa group (taxa 1169, with estimated correlation of -0.227). All the others are positive. 

Histogram of correlations between distance absolute differenced of respective correlations obtained.
```{r histogram of correlations, echo=FALSE, message=FALSE, cache = TRUE}
pairs_data_sign_corr %>% 
   ggplot(aes(x = estimate)) +
   geom_histogram(binwidth = .02, aes(y = ..count../sum(..count..))) +
   geom_vline(xintercept = mean_corr_per_taxa, color = "red") +
   labs(y = "Frequency", x = "Correlation") +
   annotate("text",
            x = mean_corr_per_taxa,
            y = .01,
            label = "Mean correlation \n of all groups' \n correlations.",
            col = "white") +
   geom_vline(xintercept = mean_corr_of_signif_levels$mean, color = "green") +
   annotate("text",
            x = mean_corr_of_signif_levels$mean,
            y = 0.03,
            label = "Mean \n of significative\n groups", 
            col = "light grey")
```

We now bind together all the pairwise tables to get a global correlation.

```{r bind big table, cache = TRUE}
all_pairs <- map_dfr(1:nrow(pairs_data_perms_props), ~{
   
   current <- slice(.data = pairs_data_perms_props, .)
   
   with(current,
        data[[1]] %>% 
           mutate(taxa = taxa,
                  prop = prop,
                  estimate = estimate,
                  p_val = p_val,
                  nb_syns = nb_syns,
                  nb_pairs = nb_pairs)
   )
})

all_taxa_cor <- with(all_pairs, cor.test(Abs_diff, Distance)) %T>% print()
```

Quick check of link between nb of pairs and proportion obtained.

Proportion of significative correlation per nb of syns in group.
```{r nb syns distribution histogram, cache= TRUE}
ggplot(pairs_data_perms_props) +
   aes(x = nb_syns) +
   geom_histogram(binwidth = 1)
```

```{r plot sign corr per nb of syns in group,cache = TRUE}
ggplot(pairs_data_perms_props) +
   aes(x = nb_syns, y = abs(estimate)) +
   geom_point() +
   geom_smooth(method = "auto", se = FALSE) +
   labs(x = "Number of SYNECHO in correlation group",
        y = "Absolute value of the correlation estimate")
```

The relation between nb of pairs (through the number of syns in the group) and the significance raises question.
First, there's only `r n_distinct(all_pairs$Distance)` different Distance values between each of the `r all_pairs %>% distinct(OTU_1, OTU_2) %>% nrow()` different pairs of SYNECHO. The min distance of which is `r min(all_pairs$Distance)`. Let's see their distribution.


```{r Distance values distribution, cache = TRUE}
all_pairs %>% 
   count(Distance) %>% 
   ggplot() +
   aes(x = Distance,
       y = n) +
   geom_col()
```

```{r}
all_pairs %>% 
   ggplot() +
   aes(y = Abs_diff, x = Distance) +
   geom_violin()
```


